Hydra Battles in Coq

### 2020-11-24 doc/part-hydras.tex
creation of a theories/ordinals/ directory

\documentclass[a4paper]{book}
\usepackage{alltt}
\usepackage{makeidx}
\usepackage{hyperref}
\usepackage{url}
\usepackage{mdframed}
\usepackage{amsmath}
\usepackage{varioref}
\usepackage{synttree, proof, centernot}
\usepackage[firstpage]{draftwatermark}
\usepackage{verbatim}
\usepackage[note]{marginote}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{dsfont}
\usepackage{threeparttable}
\setcounter{tocdepth}{1}
\input{coqmacros}

\author{Pierre Castéran\\ LaBRI, Univ. Bordeaux, CNRS UMR 5800, Bordeaux-INP\\with contributions by \'Evelyne Contejean,  Florian Hatat and Pascal Manoury}
\date{\today}
\title{Hydra Battles in Coq}
\makeindex
\begin{document}
\maketitle

\cleardoublepage

% \mbox{~}
% \clearpage
% \newpage
% \mbox{~}
\vspace{10cm}
\begin{quote} 
{ \Large {\it
``I start from one point and go as far as possible.'' }

John Coltrane.}
\end{quote}

% \begin{verse}
% Hydra hydra hydram \\

% Hydrae hydrae hydra \\

% Hydrae hydrae hydras \\

% Hydrarum hydris hydris\\
% \end{verse}

\tableofcontents

\chapter{Introduction}

\vspace{16pt}

Proof assistants are excellent tools for exploring the structure of mathematical proofs,
studying  which hypotheses are really needed, and which proof patterns are useful and/or
necessary. Since the development of a theory is represented as a bunch of computer files,
everyone is able to read the proofs with an arbitrary level of detail, or to play with the theory by writing alternate proofs or definitions.

Among all the theorems proved with the help of proof assistants like \coq{}, \isabelle{}, \hol{}, etc.,
several statements and proofs  share some interesting features:
\begin{itemize}
\item Their statements are easy to understand, even by non-mathematicians
\item Their proof requires some non-trivial mathematical tools
\item Their mechanization on computer presents some methodological interest.
\end{itemize}

This is obviously the case of the four-color theorem~\cite{fourcolors}  and the Kepler conjecture~\cite{flyspeck2015}. We do not mention impressive works like the proof of the odd-order theorem ~\cite{oddorderthm}, since understanding its statement requires a quite good mathematical culture.

Hydra games (a.k.a. \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians:
L. Kirby and J. Paris~\cite{KP82}: \emph{Accessible Independence Results for Peano Arithmetic}. 
Although the mathematical contents of this 
paper are quite advanced, the rules of hydra battles are very easy to understand. There are now several sites on Internet where you can find tutorials on hydra games, together with simulators you can play with. See, for instance, the page written by Andrej Bauer~\cite{bauer2008}.

Hydra battles, as well as Goodstein Sequences~\cite{goodstein_1944, KP82}
are a nice way to present complex termination problems.
The article by Kirby and Paris presents a proof of termination
based on ordinal numbers, as well as a proof that this termination is not
provable in Peano arithmetic. In the book dedicated to 
J.P. ~Jouannaud \cite{HommageJPJ}, N.~Dershowitz and G.~Moser  give a thorough survey on this topic~\cite{Dershowitz2007}.

Here, we present a development  for the \coq{} proof assistant, after the work of Kirby and Paris. This formalization contains the following main parts:

@@

\section{Remarks}

\subsection{Difference from Kirby and Paris's Work}
In~\cite{KP82}, Kirby and Paris showed  that there is no proof of termination of all hydra battles in Peano Arithmetic (PA).
Since we are used to writing proofs in higher order logic, the restriction to PA was quite unnatural for us. So we chosed to prove another statement without any reference to PA, by considering a class of proofs indexed by ordinal numbers upto $\epsilon_0$.

\subsection{Trust in our Proofs?}
\label{sect:trust-in-proofs}

Unlike mathematical literature, where definitions and proofs are spread over many articles and books,
the whole proof is now inside your computer. It is composed of the \texttt{.v} files you downloaded and 
parts of \coq's standard library. Thus, there is no ambiguity in our definitions and the premises of the theorems. Furthermore, you will be able to navigate through the development, using your favourite text editor or IDE, and some commands like \texttt{Search}, \texttt{Locate},  etc.

\subsection{Assumed Redundancy}

It may often happen that several definitions of a given concept, or several proofs of a given theorem are possible. If all the versions present some interest, we will make them available, since each one may be of some methodological 
interest (by illustrating some tactic of proof pattern, for instance).
We may use \coq's module system to make several proofs of a given theorem co-exist in our libraries (see also Sect~\vref{sect:alt-proofs}).
After some discussions of the pros and cons of each solution, we develop only one of them, leaving the others  as exercises or projects (i.e., big or difficult exercises).
In order to discuss which assumptions are really needed for proving a theorem, we will also present 
several aborted proofs.
Of course, do not hesitate to contribute nice proofs or alternative definitions !

It may also happen that some proof looks to be useless, because the proven theorem is a trivial consequence of another (proven too) result.
For instance, let us consider the three following statements:
\begin{enumerate}
\item There is no measure into $\mathbb{N}$ for proving the termination of all hydra battles (Sect~\vref{omega-case}).
\item There is no measure into $[0,\omega^2)$ for proving the termination of all hydra battle (Sect~\vref{omega2-case}).
\item There is no measure into $[0,\mu)$ for proving the termination of all hydra battles, for any $\mu<\epsilon_0$ (Sect\vref{sec:free-battles-case}).
\end{enumerate}

Obviously, the third theorem implies the second one, which implies the first one. So, theoretically, a library would contain only a proof of $(3)$ and remarks for $(2)$ and $(1)$. But we found it interesting to make all the three proofs available, allowing the reader to compare their common structure and notice their technical differences.
In particular, the proof of $(3)$ uses several non-trivial combinatorial properties of ordinal numbers up to $\epsilon_0$~\cite{KS81}, whilst $(1)$ and $(2)$ use simple properties of $\mathbb{N}$ and $\mathbb{N}^2$.

\subsection{About Logic}

Most of the proofs we present are \emph{constructive}. Whenever possible, we provide the user with an associated function, which she or he can apply in \gallina{} or \ocaml{} in order to get a ``concrete'' feeling of the meaning of the considered theorem.
For instance, in Chapter~\vref{chap:ketonen}, the notion of \emph{limit ordinal} is
made more ``concrete'' thanks to a function \texttt{canon} which computes every item of a sequence which converges on a given limit ordinal $\alpha$. This simply typed function allows the user/reader to make her/his own experimentations.
For instance, one can very easily compute the $42$-nd item of a sequence which converges towards $\omega^{\omega^\omega}$.

Except in the \texttt{Schutte} library, dedicated to an axiomatic presentation of the set of countable ordinal numbers, all our development is axiom-free, and respects the rules of intuitionistic logic. Note that we also use the \texttt{Equations} plug-in~\cite{sozeau:hal-01671777} in the definitition of  several rapidly growing hierarchy of functions, in Chap.~\ref{chap:alpha-large}. This plug-in imports several known-as-harmless  axioms.

% \begin{Coqsrc}
% FunctionalExtensionality.functional_extensionality_dep : 
% forall (A : Type) (B : A -> Type) (f g : forall x : A, B x),
% (forall x : A, f x = g x) -> f = g
% \end{Coqsrc}

\index{Coq!Commands!Print Assumptions}

At any place of our development, you may use the  \texttt{Print Assumptions {\it ident}} command in order to verify on which axiom the theorem {\it ident} may depend. The following example is extracted from 
Library~\href{../theories/html/hydras.Epsilon0.F_alpha.html}{hydras.Epsilon0.F\_alpha}, where we use the \texttt{coq-equations} plug-in (see Sect.~\vref{sect:wainer}).

\begin{Coqsrc}
About F_zero_eqn.
\end{Coqsrc}

\begin{Coqanswer}
F_zero_eqn : forall i : nat, F_ Zero i = S i
\end{Coqanswer}

\begin{Coqsrc}
Print Assumptions F_zero_eqn. 
\end{Coqsrc}

\begin{Coqanswer}
Axioms:
FunctionalExtensionality.functional_extensionality_dep
  : forall (A : Type) (B : A -> Type) (f g : forall x : A, B x),
    (forall x : A, f x = g x) -> f = g
Eqdep.Eq_rect_eq.eq_rect_eq
  : forall (U : Type) (p : U) (Q : U -> Type) (x : Q p) (h : p = p),
    x = eq_rect p Q x p h
\end{Coqanswer}

\subsection{Main References}

\paragraph*{Warning:}

This document is \emph{not} an introductory text for \coq{}, and there are many aspects of this proof assistant that are not covered. 
 The reader should already have some basic experience with the \coq{} system. The Reference Manual and several tutorials are available on \coq{} page~\cite{Coq}.  First chapters of textbooks like \emph{Interactive Theorem Proving and Program Development}~\cite{BC04}, \emph{Software Foundations}~\cite{SF} or  \emph{Certified Programming with Dependent Types} ~\cite{chlipalacpdt2011} will give you the right background.

\subsection{State of the Development}
The \coq{} scripts herein are in constant development since our contribution~\cite{CantorContrib} on  notations for the ordinals $\epsilon_0$ and $\Gamma_0$.
We added new material : axiomatic definition of countable ordinals after Schütte~\cite{schutte}, combinatorial aspects of $\epsilon_0$, after Ketonen and Solovay~\cite{KS81} and Kirby and Paris~\cite{KP82}, recent \coq{} technology: type classes, equations, etc.

@@

\subsection{Contributions}
Many thanks to \'Evelyne Contejean,  Florian Hatat and Pascal Manoury  for their contribution. \'Evelyne contributed libraries on the recursive path ordering (\emph{rpo}) for proving the well-foundedness of our representation of $\epsilon_0$ and $\Gamma_0$.
Florian Hatat proved many useful lemmas on countable sets, which we used in our adaptation of Schütte's formalization of countable ordinals. Pascal Manoury is integrating the ordinal $\omega^\omega$ into our hierarchy of ordinal notations.

\label{sec:orgheadline2}

Any form of contribution  is welcome: correction of errors (typos and more serious mistakes), improvement of
Coq scripts, proposition of inclusion of new chapters, and generally any
comment or proposition that would help us. The text contains several \emph{projects} which, when completed, may improve the present work.
Please do not hesitate to bring your contribution!

\subsection{Typographical Conventions}

Quotations from our  \coq{} source are displayed as follows:

  \begin{Coqsrc}
 Require Import Arith.

 Definition square (n:nat) := n * n.

 Lemma square_double : exists n:nat, n + n = square n.
 Proof.
    exists 2. 
  \end{Coqsrc}

Answers from \coq{} (including sub-goals, error messages, etc.) are displayed in slanted style
with a different background color.

 \begin{Coqanswer}
 1 subgoal, subgoal 1 (ID 5)

  ============================
   2 + 2 = square 2

 \begin{Coqsrc}
   reflexivity.
Qed.
 \end{Coqsrc}

\subsection{Active Links}
The  links which appear in this pdf  document lead are of three possible kinds of destination:
\begin{itemize}
\item Local links to the document itself,
\item External links, mainly to \coq's page,
\item Local links to pages generated by \texttt{coqdoc}. According to the current makefile (through the commands \texttt{make html} and \texttt{make pdf}), 
  we assume that the page generated from a library \texttt{XXX/YYY.v} is stored as
the relative address \texttt{../theories/html/hydras.XXX.YYY.html} (from the location of the pdf)
Thus,  active links towards our \coq{} modules may be incorrect if you got this \texttt{pdf} document otherwise than by compiling the distribution available in
\url{https://github.com/coq-community/hydra-battles}.

\subsection{Alternative or Bad Definitions}
\label{sect:alt-proofs}
Finally, we decided to include definitions or lemma statements, as well as tactics,  that lead to
dead-ends or too complex developments, with the following coloring.
Bad definitions 
 and encapsulation in modules called \texttt{Bad}, \texttt{Bad1}, etc.

\begin{Coqbad}
Require Import Lia.

Module Bad.

Definition double (n:nat)  := n + 2.

Lemma lt_double : forall n:nat, n < double  n.
Proof.
   unfold double; lia.
Qed.

Likewise, alternative, but still unexplored definitions will be presented in modules
\texttt{Alt}, \texttt{Alt1}, etc. Using these definitions is left as an implicit exercise.

\begin{Coqalt}
Require Import Arith Lia Iterates.
Module Alt.
  Definition double (n : nat) := iterate S n n.
End Alt.
\end{Coqalt}

\begin{Coqsrc}
Lemma alt_double_ok n : Nat.double n = Alt.double n.
Proof.
  unfold Alt.double, Nat.double; induction n; cbn.
  -  trivial.
  -  rewrite <- iterate_rw, iterate_S_eqn, <- IHn; lia.
Qed.
\end{Coqsrc}

\section{Acknowledgements}
\label{sec:orgheadline5}
    Many thanks to David Ilcinkas, Sylvain Salvati, Alan Schmitt and Théo Zimmerman for their help on the elaboration of this document, and to the
 members of the \emph{Formal Methods} team at laBRI for their helpful comments 
on an oral presentation of this work.

Many thanks also to the Coq development team, Yves Bertot, and members of the \emph{Coq Club} for interesting discussions about the \coq{} system and the Calculus of Inductive Constructions.

The author of the present document wishes to express his gratitude to the late Patrick Dehornoy, whose talk  was determinant for our desire to work on this topic.

I owe my interest in discrete mathematics and their relation to formal proofs and functional programming  to Srecko Brlek.  Equally, there is W. H. Burge's book ``\emph{Recursive Programming Techniques}'' ~\cite{burge} which was a great  source of inspiration.

\section{How to Install the Libraries}
\label{sec:orgheadline4}
\begin{itemize}
\item The present distribution has been checked with version 8.11.0 of the Coq proof assistant, with the plug-ins \texttt{coq-paramcoq} and \texttt{coq-equations}.

\item Just go into the root directory, and type "\texttt{make}"
\end{itemize}

\section{Comments on Exercises and Projects}

Although we do not plan to include complete solutions to the exercises, 
we think it would be useful to include comments and hints, and questions/answers from the users. In constrast, ``projects'' are supposed, once completed, to be included in the repository.

% \subsubsection{Links to the Coq source}

\section{Future Work (projects)}
\index{Projects}

@@

%-----------------------------------------------------------

\chapter{Hydras and Hydra Games}
\label{sec:orgheadline91}

@@

Technically, a \emph{hydra} is just a finite ordered tree, each node of which 
has any number of sons. Note that, contrary to the computer science tradition, we will show the hydras 
with the heads up and the foot (i.e., the root of the tree) down.
Fig.~\ref{fig:Hy} represents such  a hydra, which will be referred to as \texttt{Hy} in our examples (please look at the 
module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}).

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#Hydra}{Hydra.Hydra\_Definitions}}

@@

\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#head}{Hydra.Hydra\_Definitions}}

@@

\vspace{4mm}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html\#Hy}{Hydra.Hydra\_Examples}}

@@

\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#hcons_mult}{Hydra.Hydra\_Definitions}}

@@

\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}

@@

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}

@@

\vspace{4pt}\emph{From Module\href{../theories/html/hydras.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}

@@

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#S1}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#R2}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}.

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#round_n}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}
\label{types:Battle}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#free}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#standard}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#fight}{Hydra.Hydra\_Definitions}}

@@

% \vspace{4pt}
% \emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Lemmas.html}{Hydra.Hydra\_Lemmas}}

@@

The complete proof is in file \url{../theories/html/hydras.Hydra.BigBattle.html}.

@@

We can make bigger steps through iterations of \texttt{next}.
The functional \texttt{iterate}, similar to Standard Library's \texttt{Nat.iter},
is defined and studied in~\href{../theories/html/hydras.Prelude.Iterates.html\#iterate}{Prelude.Iterates}.

@@

\begin{Coqsrc}
Definition exp2 n := iterate (fun n => 2 * n) n 1.
\end{Coqsrc}
Using some facts (proven in 
\href{../theories/html/hydras.Hydra.BigBattle.html}{hydras.Hydra.BigBattle}),we get several  minorations.

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#Alive}{Hydra.Hydra\_Definitions}}

@@

The theorems \texttt{Alive\_free} and \texttt{Alive\_standard} of the module 
\url{../theories/html/hydras.Hydra.Hydra_Theorems.html} show that the classes \texttt{free} and \texttt{standard} satisfy this property.

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Lemmas.html\#next_round_dec}{Hydra.Hydra\_Lemmas}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Hydra.Hydra_Definitions.html\#Hvariant}{Hydra.Hydra\_Definitions}}

@@

Let us present the main steps of that proof, the script of which  is in the module ~\href{../theories/html/hydras.Hydra.Omega_Small.html}{Hydra/Omega\_Small.v} \footnote{ The name of this file means ``the ordinal $\omega$ is too small for proving the termination of [free] hydra battles ''. In effect, the elements of $\omega$, considered as a set, are just the natural numbers (see next chapter for more details)}.

@@

More intuitively, let us have a look at Figure~\ref{fig:ordinal-sequence}. It presents a small sequence of ordinal numbers, which extends the sequence of natural numbers.

@@

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for an infinite sequence of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, between $17$ and $\omega$.
\item Each ordinal printed in black is the immediate successor of another ordinal. We call it a \emph{successor} ordinal. For instance, $12$ is the successor of $11$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals (displayed in red)  that  follow immediately dots are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). It is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item Any ordinal is  either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

\subsection{Well-ordered Sets}
Let us start with some definitions.
A  \emph{well-ordered set} is a set provided with a binary relation $<$ which has the following properties.
\begin{description}
\item[irreflexivity] : $\forall x\in A, x\not< x$
\item[transitivity] : $\forall x\,y\,z\in A, x<y \Rightarrow y<z \Rightarrow x<z$
\item[trichotomy]: $\forall x\,y\in A, x<y \vee x = y \vee y < x$
\item[well foundedness]: $<$ is well-founded (every element of $A$ is accessible)\footnote{In classical mathematics, we would say that there is no infinite sequence $a_1>a_2> \dots a_n> a_{n+1}\dots$ in $A$. Please refer to any documenation on \coq{} for having more details on well-foundedness and accessibility.}.

\end{description}

@@

\begin{align*}
(i,j) < (i,k) & \;\textit{if} \; j < k\\
(0,k) < (1,l) & \;\textit{for\,any}\;k \;\textit{and} \; l
\end{align*}

@@

In our adaptation to \coq{}, we declare a type \texttt{Ord}, a binary relation \texttt{lt} (with infix notation \texttt{"\_<\_"}, and assume Schütte's axiom. In Chapter~\ref{chap:schutte},
we derive some interesting properties of countable ordinals from these axioms.
It is interesting to compare proofs of a given property (for instance the associativity of addition) in the computatuinal framework of some ordinal notation, and in the axiomatic model of Schütte.

@@

\vspace{4pt}
\noindent\emph{From Library~\href{../theories/html/hydras.OrdinalNotations.Definitions.html}{OrdinalNotations.Definitions}}

@@

\vspace{4pt}
\noindent\emph{From Library~\href{../theories/html/hydras.OrdinalNotations/ON_Omega.html}{OrdinalNotations.ON\_Omega}}

@@

\vspace{4pt}
\noindent\emph{From Library~\href{../theories/html/hydras.OrdinalNotations/ON_plus.html}{OrdinalNotations.ON\_plus}}

@@

\vspace{4pt}
\noindent\emph{From Module~\href{../theories/html/hydras.OrdinalNotations.ON_Omega_plus_omega.html}{OrdinalNotations.ON\_Omega\_plus\_omega}}

@@

The following definitions are in Library \href{../theories/html/hydras.Prelude.MoreOrders.html}{Prelude.MoreOrders}.

@@

\noindent\emph{From Module~\href{../theories/html/hydras.OrdinalNotations.ON_Omega_plus_omega.html}{OrdinalNotations.ON\_Omega\_plus\_omega}}

@@

\noindent\emph{From Module~\href{../theories/html/hydras.OrdinalNotations.Generic.html}{OrdinalNotations.Generic}}
\begin{Coqsrc}
  Definition ZeroLimitSucc_dec {A:Type}{lt: relation A}
           {compare : A -> A -> comparison}
           {on : ON lt compare} :=
  forall alpha,
    {Least alpha} +
    {Limit alpha} +
    {beta: A | Successor alpha beta}.
(* ... *)
\end{Coqsrc}

>>>>

\section*{Introduction}

In this part, we present a development  for the \coq{} proof assistant, after the work of Kirby and Paris. This formalization contains the following main parts:

@@

%\section{Remarks}

\subsection*{Difference from Kirby and Paris's Work}
In~\cite{KP82}, Kirby and Paris showed  that there is no proof of termination of all hydra battles in Peano Arithmetic (PA).
Since we are used to writing proofs in higher order logic, the restriction to PA was quite unnatural for us. So we chosed to prove another statement without any reference to PA, by considering a class of proofs indexed by ordinal numbers upto $\epsilon_0$.

\subsection*{State of the Development}
The \coq{} scripts herein are in constant development since our contribution~\cite{CantorContrib} on  notations for the ordinals $\epsilon_0$ and $\Gamma_0$.
We added new material : axiomatic definition of countable ordinals after Schütte~\cite{schutte}, combinatorial aspects of $\epsilon_0$, after Ketonen and Solovay~\cite{KS81} and Kirby and Paris~\cite{KP82}, recent \coq{} technology: type classes, equations, etc.

@@

\section*{Future Work (projects)}
\index{Projects}

@@

\subsection*{Main References}

\chapter{Hydras and Hydra Games}

\label{sec:orgheadline91}

@@

Technically, a \emph{hydra} is just a finite ordered tree, each node of which 
has any number of sons. Note that, contrary to the computer science tradition, we will show the hydras 
with the heads up and the foot (i.e., the root of the tree) down.
Fig.~\ref{fig:Hy} represents such  a hydra, which will be referred to as \texttt{Hy} in our examples (please look at the 
module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}).

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html\#Hydra}{Hydra.Hydra\_Definitions}}

@@

\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html\#head}{Hydra.Hydra\_Definitions}}

@@

\vspace{4mm}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Examples.html\#Hy}{Hydra.Hydra\_Examples}}

@@

\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html\#hcons_mult}{Hydra.Hydra\_Definitions}}

@@

\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}

@@

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}

@@

\vspace{4pt}\emph{From Module\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}

@@

\vspace{4pt}
\noindent
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html\#S1}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html\#R2}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}.

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html\#round_n}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html}{Hydra.Hydra\_Definitions}}
\label{types:Battle}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html\#free}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html\#standard}{Hydra.Hydra\_Definitions}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html\#fight}{Hydra.Hydra\_Definitions}}

@@

% \vspace{4pt}
% \emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Lemmas.html}{Hydra.Hydra\_Lemmas}}

@@

The complete proof is in file \url{../theories/html/hydras.ordinals.Hydra.BigBattle.html}.

@@

We can make bigger steps through iterations of \texttt{next}.
The functional \texttt{iterate}, similar to Standard Library's \texttt{Nat.iter},
is defined and studied in~\href{../theories/html/hydras.ordinals.Prelude.Iterates.html\#iterate}{Prelude.Iterates}.

@@

\begin{Coqsrc}
Definition exp2 n := iterate (fun n => 2 * n) n 1.
\end{Coqsrc}
Using some facts (proven in 
\href{../theories/html/hydras.ordinals.Hydra.BigBattle.html}{hydras.Hydra.BigBattle}),we get several  minorations.

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html\#Alive}{Hydra.Hydra\_Definitions}}

@@

The theorems \texttt{Alive\_free} and \texttt{Alive\_standard} of the module 
\url{../theories/html/hydras.ordinals.Hydra.Hydra_Theorems.html} show that the classes \texttt{free} and \texttt{standard} satisfy this property.

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Lemmas.html\#next_round_dec}{Hydra.Hydra\_Lemmas}}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.ordinals.Hydra.Hydra_Definitions.html\#Hvariant}{Hydra.Hydra\_Definitions}}

@@

Let us present the main steps of that proof, the script of which  is in the module ~\href{../theories/html/hydras.ordinals.Hydra.Omega_Small.html}{Hydra/Omega\_Small.v} \footnote{ The name of this file means ``the ordinal $\omega$ is too small for proving the termination of [free] hydra battles ''. In effect, the elements of $\omega$, considered as a set, are just the natural numbers (see next chapter for more details)}.

@@

For the computer scientist, ordinals are tools for proving the totality of a given recursive function, or termination of a transition system. \emph{Ordinal arithmetic} 
provides a set of functions whose properties, like \emph{monotony}, allow to define \emph{variants}, \emph{i.e.} strictly decreasing measures used in proofs of termination.

Let us have a look at Figure~\ref{fig:ordinal-sequence}. It presents a small sequence of ordinal numbers, which extends the sequence of natural numbers.

@@

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for  infinite sequences of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, between $17$ and $\omega$.
\item Each ordinal printed in black is the immediate successor of another ordinal. We call it a \emph{successor} ordinal. For instance, $12$ is the successor of $11$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals (displayed in red)  that  follow immediately dots are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). It is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item Any ordinal is  either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

\subsection{Well-ordered Sets}
Let us start with some definitions.
A  \emph{well-ordered set} is a set provided with a binary relation $<$ which has the following properties.
\begin{description}
\item[irreflexivity] : $\forall x\in A, x\not< x$
\item[transitivity] : $\forall x\,y\,z\in A, x<y \Rightarrow y<z \Rightarrow x<z$
\item[trichotomy]: $\forall x\,y\in A, x<y \vee x = y \vee y < x$
\item[well foundedness]: $<$ is well-founded (every element of $A$ is accessible)\footnote{In classical mathematics, we would say that there is no infinite sequence $a_1>a_2> \dots a_n> a_{n+1}\dots$ in $A$. In contrast, \coq's standard library contains
an inductive definition of a predicate \texttt{Acc} which allows us to write 
constructive proofs of accessibility (See \href{https://coq.inria.fr/distrib/current/stdlib/Coq.Init.Wf.html}{Coq.Init.Wf}).}.
\end{description}

@@

\begin{align*}
(i,j) < (i,k) & \;\textit{\textbf{if} }\; j < k\\
(0,k) < (1,l) & \;\textit{\textbf{for\,any}}\;k \;\textit{\textbf{and}} \; l
\end{align*}

@@

In our adaptation to \coq{}, we declare a type \texttt{Ord}, a binary relation \texttt{lt} (with infix notation \texttt{"\_<\_"}, and assume Schütte's axiom. In Chapter~\ref{chap:schutte},
we derive some interesting properties of countable ordinals from these axioms.
It is interesting to compare proofs of a given property (for instance the associativity of addition) in the computational framework of some ordinal notation, and in the axiomatic model of Schütte.

@@

\vspace{4pt}
\noindent\emph{From Library~\href{../theories/html/hydras.ordinals.OrdinalNotations.Definitions.html}{OrdinalNotations.Definitions}}

@@

\vspace{4pt}
\noindent\emph{From Library~\href{../theories/html/hydras.ordinals.OrdinalNotations/ON_Omega.html}{OrdinalNotations.ON\_Omega}}

@@

\vspace{4pt}
\noindent\emph{From Library~\href{../theories/html/hydras.ordinals.OrdinalNotations/ON_plus.html}{OrdinalNotations.ON\_plus}}

@@

\vspace{4pt}
\noindent\emph{From Module~\href{../theories/html/hydras.ordinals.OrdinalNotations.ON_Omega_plus_omega.html}{OrdinalNotations.ON\_Omega\_plus\_omega}}

@@

The following definitions are in Library \href{../theories/html/hydras.ordinals.Prelude.MoreOrders.html}{Prelude.MoreOrders}.

@@

\noindent\emph{From Module~\href{../theories/html/hydras.ordinals.OrdinalNotations.ON_Omega_plus_omega.html}{OrdinalNotations.ON\_Omega\_plus\_omega}}

@@

\noindent\emph{From Module~\href{../theories/html/hydras.ordinals.OrdinalNotations.Generic.html}{OrdinalNotations.Generic}}
\begin{Coqsrc}
  Definition ZeroLimitSucc_dec {A:Type}{lt: relation A}
           {compare : A -> A -> comparison}
           {on : ON lt compare} :=
  forall alpha,
    {Least alpha} +
    {Limit alpha} +
    {beta: A | Successor alpha beta}.
(* ... *)
\end{Coqsrc}

\noindent\emph{From Module~\href{../theories/html/hydras.ordinals.OrdinalNotations.ON_Omega_plus_omega.html}{OrdinalNotations.ON\_Omega\_plus\_omega}}
